Algebraic File Synchronization: Adequacy and Completeness

### 2018-07-05 p2/algebra.tex
Towards a new proof
Output labels for easier reviewing
Process todos

\section{Towards a Formal Algebra}\label{sec_algebra}

Similarly to \cite{NREC} we can consider creating an algebra over command sequences
which will enable us to draw conclusions about the behaviors of the sequences
independently of filesystems.
This would provide a secondary model of (filesystem) commands above the model a filesystems
defined in this paper. The new model would no longer describe filesystems as such,
but use known relationships between sequences of commands as its starting point.

In this algebra, equivalence ($\equiv$) and extension ($\eqext$) would become algebraic relations
between sequences, and logical rules involving them
(e.g. $ A\equiv B \Rightarrow S\cc A\cc T\equiv S\cc B\cc T $) would be re-cast as inference rules.
In addition to these, the \namecrefs{ax_separate_commute} listed in \cref{section_axioms}
would become axioms that we accept as true, and from which other true statements can be derived
using the inference rules.
Such a system can allow us to deduce, for example, whether two sequences are equivalent
or one extends the other
without reference to filesystems.

We note, however, that the \namecrefs{ax_separate_commute} to be used as axioms 
specify the relationships between the nodes
the commands act on, which would require a more complicated set of symbols to represent
in the algebra.
To avoid this, we think it will be more fruitful to regard commands (and sequences)
on different nodes as different, and, in effect, have nine times as many commands 
in the algebra as there are nodes in the namespace of the filesystems.
This allows us to reduce the number of types of symbols in our algebra, and regard
the \namecrefs{ax_separate_commute} not as axioms, but as templates for axioms
from which all axioms (for pairs of separate nodes, etc.) can be generated.

We expect that the soundness and completeness of such an algebra can be proven
similarly to the proofs described in \cite{NREC}.
Indeed, results in the current paper can serve as building blocks of such a proof:
the computer program used to check the \namecrefs{ax_separate_commute}
in effect proves the soundness of all axioms,
and \cref{simple_reorder_equiv} proves completeness in a limited sense.
It is also worth noting that the majority of proofs presented in this paper
do not actually refer to specific filesystems or the definition of commands,
but draw on known relationships and the
\namecrefs{ax_separate_commute} in \cref{section_axioms}.
In other words, many proofs are transferable to the algebra that we describe here.

In a more complete algebra from which the correctness and completeness
of reconciliation could also be derived,
one can also include symbols, inference rules and axioms for 
$\works{\seqset{A}}$ and $\worksc{\seqset{A}}{\seqset{B}}$.
We see proving the soundness and completeness of this extended algebra
an intriguing problem that is worthy of further research.

>>>>

\section{Towards a Filesystem-Free Algebra}\mlabel{sec_algebra}

Similarly to \cite{NREC} we can consider creating an algebra over command sequences
which would enable us to draw conclusions about the behaviors of the sequences
without considering the underlying filesystems.
This would provide a secondary model of (filesystem) commands above (and indepentent of) the model of filesystems
defined in this paper. The new model would no longer describe filesystems as such,
but would use known relationships between sequences of commands as its starting point.

In this algebra, equivalence ($\equiv$) and extension ($\eqext$) would become algebraic relations
between sequences, and logical rules involving them
(e.g. $ A\equiv B \Rightarrow S\cc A\cc T\equiv S\cc B\cc T $) would be re-cast as inference rules.
In addition to these, the \namecrefs{ax_separate_commute} listed in \cref{section_axioms}
would become axioms that we accept as true, and from which other true statements can be derived
using the inference rules.
Such a system would allow us to deduce, for example, whether two sequences are equivalent
or one extends the other without reference to filesystems.

We note, however, that the \namecrefs{ax_separate_commute} to be used as axioms 
specify the relationships between the nodes
the commands act on, which would require a more complicated set of symbols to represent
in the new algebra.
To avoid this, we think it will be more fruitful to regard commands (and sequences)
on different nodes as different, and, in effect, have nine times as many commands 
in the algebra as there are nodes in the namespace of the filesystems.
This will allow us to reduce the number of types of symbols in our algebra, and regard
the \namecrefs{ax_separate_commute} not as axioms, but as templates for axioms
from which all axioms (for pairs of separate nodes, etc.) can be generated.

We expect that the soundness and completeness of such an algebra can be proven
similarly to the proofs described in \cite{NREC}.
Indeed, results in the current paper can serve as building blocks of such a proof:
derivations in \cref{axiom_proof} for the \namecrefs{ax_separate_commute}
in effect proves the soundness of all axioms,
and \cref{simple_reorder_equiv} proves completeness in a limited sense.
It is also worth noting that the majority of proofs presented in this paper
do not actually refer to specific filesystems or the definition of commands,
but draw on known relationships between the commands themselves.
In other words, many proofs are transferable to the algebra that we describe here.

In a more complete algebra from which the correctness and completeness
of reconciliation could also be derived,
one would also include symbols, inference rules and axioms for 
$\worksmnil{\seqset{A}}$ and $\worksm{\seqset{A}}{\seqset{B}}$.
We see proving the soundness and completeness of this extended algebra
an intriguing problem that is worthy of further research.

### 2018-07-05 p2/axiom-proof.tex
Restating rules for proofs
Process todos

TODO From filesystem definition

In this section we offer proofs for the \namecrefs{ax_separate_commute} listed in
\cref{rules_lemma}.
For the purposes of the proofs we extend the notion of filesystems 
to the full set of functions mapping $\setn$ to $\setv$, even if they
do not have the tree property.
In such cases we continue to say that the filsystem is broken,
but we also continue to describe the filesystem values at nodes in $\setn$.
Also, in order to pinpoint the region that violates the tree property,
we say that the filesystem is \emph{broken at $n$} if 
$n$ is not a dictionary, but has at least one non-empty child.

Accordingly, where needed, we also implicitly extend filesystem commands
so that they would return a filesystem even if it violates the tree property.
They continue to be undefined in cases where their input types are incompatible
with the input filesystem.
We note that once a filesystem is broken, it stays broken, even if
after a subsequent command it would no longer violate the tree property.

@@

\cref{ax_separate_commute}

@@

\cref{ax_separate_nobreaks}

@@

\cref{ax_same_breaks}

@@

\cref{ax_same_emptyseq}

@@

\cref{ax_same_singlec}

@@

\cref{ax_directchild_breaks}

@@

\cref{ax_directparent_breaks}

@@

\cref{ax_distantrel_breaks}

@@

\cref{ax_child_assert}

@@

\cref{ax_parent_assert}

@@

\cref{ax_assert}

>>>>

In this section we derive the \namecrefs{ax_separate_commute} listed in
\cref{rules_lemma} from our filesystem model.
For the purposes of the proofs we extend the notion of filesystems 
to the full set of functions mapping $\setn$ to $\setv$, even if they
do not have the tree property.
This is so that we can describe the values they would acquire under certain nodes,
and pinpoint where they may be violating the tree property.
If they do, we continue to say that the filesystem is broken,
and say that the filesystem is \emph{broken at $n$} if 
$n$ is not a dictionary, but has at least one non-empty child.
We continue to require that filesystems are not empty at finitely many nodes.

Accordingly, where needed, we also implicitly extend filesystem commands
so that they would return a filesystem even if it violates the tree property.
We continue to say that they break the filesystem in these cases,
and commands continue to be undefined where their input types are incompatible
with the input filesystem.
We note that once a filesystem is broken, it stays broken, even if
after a subsequent command it would no longer violate the tree property.

@@

Below we restate the \namecrefs{ax_separate_commute} defined in
\cref{rules_lemma}.

\begin{myaxproof}
% ax_separate_commute
\axaxseparatecommute
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_separate_nobreaks}
\axaxseparatenobreaks
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_same_breaks}
\axaxsamebreaks
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_same_emptyseq}
\axaxsameemptyseq
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_same_singlec}
\axaxsamesinglec
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_directchild_breaks}
\axaxdirectchildbreaks
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_directparent_breaks}
\axaxdirectparentbreaks
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_distantrel_breaks}
\axaxdistantrelbreaks
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_child_assert}
\axaxchildassert
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_parent_assert}
\axaxparentassert
\end{myaxproof}

@@

\begin{myaxproof}
% \cref{ax_assert}
\axaxassert
\end{myaxproof}

### 2018-07-05 p2/conclusion.tex
Improved introduction
Process todos

\section{Conclusions and Further Research}

Apart from constructing an algebra,
there are many other ways in which further work can extend the current results.
An important extension would be to
consider reconciling not only two, but more replicas in a single step and
prove the correctness and maximality of the algorithm proposed,
or show that it is impossible to satisfy these criteria.

A related problem is to extend the system and the proofs
to allow for cases where reconciliation cannot
complete fully
or if only a subset of the replicas are reconciled 
(e.g. due to network partitioning),
both of which would result in a state where different replicas
have different common ancestors, that is,
the updates specific to the replicas start from different points
in the update history of the filesystem.
Existing research can offer pointers as to how such cases can be modelled
in our algebraic system.
Parker et al. \cite{PPRS} and Cox and Josephson \cite{CJ}
describe version vectors (update histories) kept as metadata,
while Chong and Hamadi present distributed algorithms that allow incremental synchronization \cite{CH}.
Representing individual updates to files
in their modification histories (as described in \cite{CJ})
as separate commands could also enable an algebraic synchronizer to reconcile otherwise
conflicting updates and resolve partial reconciliations.

An extension of our results could also entail relaxing
the simplifying assumptions we used.
Identifying which results depend on the assumption that $|\setd|=1$
and devising updated proofs and algorithms for when $|\setd|>1$,
or specifying a pre-processing step that reconciles meta-information in directories
can enable the synchronizer
to identify and resolve a wider range of conflicts.
Similarly, designing pre- and post-processors that
present removing and creating a node
not as separate updates but as a single rename (move),
could contribute to the user-friendliness of the reconciler
and help avoid human error.

And finally, we hope that this work, together with \cite{NREC}, provides
a blueprint of constructing an algebra of commands for different storage protocols
(e.g. XML trees, mailbox folders, generic relational databases, etc.),
and of demonstrating the adequacy and completeness of update and conflict detection and reconciliation
algorithms defined over it.
This, in turn, can offer formal verification of the algorithms underlying
specific implementations in a variety of synchronizers.
Alternatively, by generalizing the parent--child relationships between filesystem nodes,
the demonstrated properties of minimal sequences of commands
and conditional operation of sequences ($\workssign$)
may also contribute to future research into algebraic structures
constrained by predefined sparse connections between their elements.

>>>>

\section{Conclusions and Further Research}\mlabel{sec_conclusion}

Apart from constructing a filesystem-free algebra,
there are many other ways in which further work can extend the current results.
An important extension would be to
consider reconciling not only two, but more replicas in a single step and
prove the correctness and maximality of the algorithm proposed,
or show that it is impossible to satisfy these criteria.

A related problem is to extend the system and the proofs
to allow for cases where reconciliation cannot
complete fully
or if only a subset of the replicas are reconciled 
(e.g. due to network partitioning),
both of which would result in a state where different replicas
have different common ancestors, that is,
the updates specific to the replicas start from different points
in the update history of the filesystems.
Existing research can offer pointers as to how such cases can be modelled
in our algebraic system.
Parker et al. \cite{PPRS} and Cox and Josephson \cite{CJ}
describe version vectors (update histories) kept as metadata,
while Chong and Hamadi present distributed algorithms that allow incremental synchronization \cite{CH}.
Representing individual updates to files
in their modification histories (as described in \cite{CJ})
as separate commands could also enable an algebraic synchronizer to reconcile otherwise
conflicting updates and resolve partial reconciliations.

An extension of our results could also entail relaxing
the simplifying assumptions we used.
% TODO TODO TODO
Identifying which results depend on the assumption that $|\setd|=1$
and devising updated proofs and algorithms for when $|\setd|>1$,
or specifying a pre-processing step that reconciles meta-information in directories
can enable the synchronizer
to identify and resolve a wider range of conflicts.

Similarly, designing pre- and post-processors that
present removing and creating a node
not as separate updates but as a single rename (move),
could contribute to the user-friendliness of the reconciler
and help avoid human error.

And finally, we hope that this work, together with \cite{NREC}, provides
a blueprint of constructing an algebra of commands for different storage protocols
(e.g. XML trees, mailbox folders, generic relational databases, etc.),
and of demonstrating the adequacy and completeness of update and conflict detection and reconciliation
algorithms defined over it.
This, in turn, can offer formal verification of the algorithms underlying
specific implementations in a variety of synchronizers.
Alternatively, by generalizing the parent--child relationships between filesystem nodes,
the demonstrated properties of minimal sequences of commands
and domains of sets of sequences ($\worksmeqsign$)
may also contribute to future research into algebraic structures
constrained by predefined sparse connections between their elements.

### 2018-07-05 p2/def-filesystem.tex
Proofs for the rules
Process todos

\begin{mydef}[Filesystem values: $\vald$, $\valfx$, $\empt$, $\typeeq$, $\eqclass{v}$]
The set of values $\setv$ is partitioned into files, directories, and the empty value $\empt$.
% TODO May be able to delete \typeeq
For $v_1,v_2\in\setv$ we write $v_1\typeeq v_2$ iff
$v_1$ and $v_2$ are both files, are both directories, or are both $\empt$.
As usual, we write $[v]$ for the the equivalence class of $v\in\setv$, which represents its type.
File and directory values are also denoted by $\valfx$ and $\vald$, respectively,
with or without indices.
\end{mydef}

@@

Every filesystem has a so-called \emph{tree property}, which means that
if the filesystem is not empty at a node, and the node has a parent,
then there must be a directory at the parent node.
Using the notation introduced above, we can formally express this as follows.
\begin{mydef}[Tree property]
A filesystem $\FS:\setn \rightarrow \setv$ has the tree property iff
\[ \forall n\in\setn
\FS(n) \neq \empt \Rightarrow \FS(\parentf{n}) \typeeq \vald \]
wherever $\parent(n)$ is defined, that is, $n$ has a parent.
\end{mydef}

@@

\myskip
In the rest of this paper,
we fix the skeleton $\langle\setn,\parent\rangle$,
and the set of values $\setv$.
$\FS$ (with or without indices) denotes a filesystem,
and $n$, $m$ and $o$ are nodes in $\setn$,
and we write $\nn$ for the parent of $n$, assuming it exists.

>>>>

\begin{mydef}[Filesystem values: $\vald$, $\valfx$, $\empt$, $\eqclass{v}$]
The set of values $\setv$ is partitioned into files, directories, and the empty value $\empt$.
As usual, we write $[v]$ for the the equivalence class of $v\in\setv$ according to this partition,
which represents its type.
File and directory values are also denoted by $\valfx$ and $\vald$, respectively,
with or without indices.
\end{mydef}

@@

Every filesystem has a so-called \emph{tree property}, which means that
if the filesystem is not empty at a node, and the node has a parent,
then there must be a directory at the parent node.
Using the notation introduced above, we can formally express this as follows.
\begin{mydef}[Tree property]
A filesystem $\FS:\setn \rightarrow \setv$ has the tree property iff
\[ \forall n\in\setn
\FS(n) \neq \empt \; \Longrightarrow \; [\FS(\parentf{n})] = [\vald] \]
wherever $\parent(n)$ is defined, that is, $n$ has a parent.
\end{mydef}

@@

\myskip
In the rest of this paper,
we fix the skeleton $\langle\setn,\parent\rangle$,
and the set of values $\setv$.
$\FS$ (with or without indices) denotes a filesystem,
and $n$, $m$ and $o$ are nodes in $\setn$.

### 2018-07-05 p2/works-inverse.tex
Process todos

\begin{myax}\mlabel{combine_independent_commands}
The combination of independent commands is defined on all filesystems
where both of the commands are defined:
\[ \alpha\indep \beta \Rightarrow \worksmbx{\alpha, \beta}{\alpha\cc \beta}. \]
\end{myax}
\begin{proof}
TODO: they are the same? FF?

We name this proposition a \namecref{combine_independent_commands} 
because to prove it, we must reach back to our filesystem model.
Let $\alpha=\cxynv$ and $\beta=\czwmv$.
The proof is by contradiction;
assume that there is a filesystem $\FS$ for which
$\cxynv\aFS\neq\fsbroken$ and $\czwmv\aFS\neq\fsbroken$, but
$(\cxynv\cc \czwmv)\aFS=\fsbroken$.
We know $\cxynv\aFS\neq\fsbroken$ so it must be applying 
$\czwmv$ that breaks it.
Applying a command can only result in a broken filesystem in two cases.
First, if the input type does not match the filesystem,
but we know $[\FS(m)]=[w]$ and so
$[\cxynv\aFS(m)]=[w]$ as based on \cref{incomparable_is_independent}, $n\neq m$.
Second, if the new filesystem violates the tree property.
This again cannot be the case because we also know that $n\unrel m$
and the tree property only depends on the types of values at the parent and children of $m$,
which therefore cannot be changed by $\cxynv$.
\end{proof}

>>>>

\begin{mylem}\mlabel{combine_independent_commands}
The combination of independent commands is defined on all filesystems
where both of the commands are defined:
\[ \alpha\indep \beta \Rightarrow \worksmbx{\alpha, \beta}{\alpha\cc \beta}. \]
\end{mylem}
\begin{proof}
TODO: they are the same? FF?

In the following proof we reach back to our filesystem model.
Let $\alpha=\cxynv$ and $\beta=\czwmv$.
The proof is by contradiction;
assume that there is a filesystem $\FS$ for which
$\cxynv\aFS\neq\fsbroken$ and $\czwmv\aFS\neq\fsbroken$, but
$(\cxynv\cc \czwmv)\aFS=\fsbroken$.
We know $\cxynv\aFS\neq\fsbroken$ so it must be applying 
$\czwmv$ that breaks it.
Applying a command can only result in a broken filesystem in two cases.
First, if the input type does not match the filesystem,
but we know $[\FS(m)]=[w]$ and so
$[\cxynv\aFS(m)]=[w]$ as based on \cref{incomparable_is_independent}, $n\neq m$.
Second, if the new filesystem violates the tree property.
This again cannot be the case because we also know that $n\unrel m$
and the tree property only depends on the types of values at the parent and children of $m$,
which therefore cannot be changed by $\cxynv$.
\end{proof}
